An Object-Logic Explanation for the Inconsistency in G\"odel's Ontological Theory (Extended Abstract)

### 2016-07-11 Papers/2016/KI/main.tex
update from Christoph

\newcommand{\verbsize}{\small}

@@

\title{An Object-Logic Explanation for the Inconsistency in G\"odel's Ontological Theory}
\author{Christoph Benzm\"uller\inst{1}\thanks{This work was supported by
    the German Research Foundation DFG grant BE2501/9-1,2} \and Bruno Woltzenlogel Paleo\inst{2} 
}

@@

\begin{abstract}
  This paper discusses the 
  inconsistency in G\"odel's
  ontological argument. Despite the popularity of G\"odel's argument, this
  inconsistency remained unnoticed until 2013, 
  when it was detected automatically by the 
  higher-order theorem prover \textsc{Leo-II} \cite{ECAI2014}. Complementing the meta-logic explanation for the inconsistency available in our IJCAI 2016 paper \cite{C55}, we present here a new purely object-logic explanation that does not rely on semantic argumentation.

@@

G\"odel's omission has been
considered inessential and merely an oversight by many. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analyses and criticisms of the
argument.
However, as explained here, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_Goedel}
where the model finder Nitpick \cite{Nitpick} confirms consistency). In
personal communication, Dana Scott confirmed that he was unaware
that G\"odel's original axioms were inconsistent.

@@

In our experiments in this branch of metaphysics we
utilise an embedding of HOMLs, such as \textbf{K}, \textbf{KB} and
\textbf{S5} with various domain conditions (possibilist and actualist
quantification), in HOL. More precisely, formulas in HOML are \emph{lifted}, i.e., converted
into predicates over worlds, which are themselves explicitly
represented as terms. The logical constants of HOML are translated to
HOL terms in such a way that, for instance,
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation,
%\cite{DBLP:journals/logcom/Ohlbach91}, 
which is here intra-logically realized --- and extended for
quantifiers --- in HOL by stating a set of equations defining the
logical constants. The resulting object logic is
the HOML \textbf{K} with rigid terms and constant domains (possibilist
quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
using an existence predicate to guard the quantifiers.

@@

In the typical workflow during an attempt to prove a conjecture with a
theorem prover, it is customary to check the consistency of the axioms
first. For if the axioms are inconsistent, anything (including the
conjecture) would be trivially derivable in classical logic (\emph{ex
  falso quodlibet}). Surprisingly, when this routine check was
performed on G\"odel's axioms \cite{C40}, the \textsc{Leo-II} prover
claimed that the axioms were inconsistent. Unfortunately, the
refutation generated by \textsc{Leo-II} was barely human-readable. The refutation was based on machine-oriented inference rules (the higher-order resolution calculus \cite{W47}), and the
text file had 153 lines (with an average of 184
  characters per line) and used a machine-oriented syntax (TPTP THF
\cite{J22}).

@@

\item $\pos \pos \bot$ \emph{is derivable:} and indeed, it can be derived (using a recently developed natural deduction calculus for modal logic \K \cite{CSR}) as follows:
%\smallskip
\begin{center}
\includegraphics[width=0.5\textwidth]{Derivation}
\end{center}
%\smallskip

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Lemma dia_box_false_to_false_meta: [(dia (box mFalse))] -> [mFalse].
\Proof. intro H. intro w.
destruct (H w) as [w0 [R0 H0]]. destruct (H w0) as [w1 [R1 H1]].
box_elim H0 w1 HF. unfold mFalse in HF. destruct HF as [p [HF1 HF2]].
contradiction. \Qed.
\end{Verbatim}

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Lemma mimplies_to_mnot: [mforall p:o, (p m-> mFalse) m-> (m~ p)].
\Proof. mv. intro p. intro H. intro H0. 
destruct (H H0) as [p0 [H1 H2]]. apply H2. exact H1. \Qed.

>>>>

\newcommand{\verbsize}{\scriptsize}

@@

\title{An Object-Logic Explanation for the Inconsistency in G\"odel's
  Ontological Theory (Extended Abstract)}
\author{Christoph Benzm\"uller\inst{1}\thanks{This work was supported by
    the German Research Foundation DFG grant BE2501/9-1,2} \and Bruno Woltzenlogel Paleo\inst{2} 
}

@@

\begin{abstract}
  This paper discusses the 
  inconsistency in G\"odel's
  ontological argument. Despite the popularity of G\"odel's argument, this
  inconsistency remained unnoticed until 2013, 
  when it was detected automatically by the 
  higher-order theorem prover \textsc{Leo-II}. Complementing the meta-logic explanation for the inconsistency available in our IJCAI 2016 paper \cite{C55}, we present here a new purely object-logic explanation that does not rely on semantic argumentation.

@@

G\"odel's omission has been
considered inessential and merely an oversight by many. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analyses and criticisms of the
argument.
However, as explained here, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_Goedel},
where the model finder Nitpick \cite{Nitpick} confirms consistency). In
personal communication, Dana Scott confirmed that he was unaware
that G\"odel's original axioms were inconsistent.

@@

In our experiments in this branch of metaphysics we
utilise an embedding of HOMLs, such as \textbf{K}, \textbf{KB} and
\textbf{S5} with various domain conditions (possibilist and actualist
quantification), in HOL. More precisely, formulas in HOML are \emph{lifted}, i.e., converted
into predicates over worlds, which are themselves explicitly
represented as terms. The logical constants of HOML are translated to
HOL terms in such a way that, for instance,
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_0$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation,
%\cite{DBLP:journals/logcom/Ohlbach91}, 
which is here intra-logically realized --- and extended for
quantifiers --- in HOL by stating a set of equations defining the
logical constants. The resulting object logic is
the HOML \textbf{K} with rigid terms and constant domains (possibilist
quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
using an existence predicate to guard the quantifiers.

@@

In the typical workflow during an attempt to prove a conjecture with a
theorem prover, it is customary to check the consistency of the axioms
first. For if the axioms are inconsistent, anything (including the
conjecture) would be trivially derivable in classical logic (\emph{ex
  falso quodlibet}). Surprisingly, when this routine check was
performed on G\"odel's axioms \cite{C40}, the \textsc{Leo-II} prover
claimed that the axioms were inconsistent. Unfortunately, the
refutation generated by \textsc{Leo-II} was barely human-readable. The
refutation was based on machine-oriented inference rules (a higher-order resolution calculus \cite{J30}), and the
text file had 153 lines (with an average of 184
  characters per line) and used a machine-oriented syntax (TPTP THF
\cite{J22}).

@@

\item $\pos \pos \bot$ \emph{is derivable:} and indeed, it can be
  derived (using a recently developed natural deduction calculus for
  modal logic \K \cite{CSR}) as follows: \\
%\smallskip

\begin{center}
\includegraphics[width=0.4\textwidth]{Derivation}
\end{center}
%\smallskip

@@

\begin{Verbatim}[frame=single,commandchars=\\\{\},fontsize=\verbsize]
\Lemma dia_box_false_to_false_meta: [(dia (box mFalse))] -> [mFalse].
\Proof. intro H. intro w.
destruct (H w) as [w0 [R0 H0]]. destruct (H w0) as [w1 [R1 H1]].
box_elim H0 w1 HF. unfold mFalse in HF. destruct HF as [p [HF1 HF2]].
contradiction. \Qed.
\end{Verbatim}

@@

\begin{Verbatim}[frame=single,commandchars=\\\{\},fontsize=\verbsize]
\Lemma mimplies_to_mnot: [mforall p:o, (p m-> mFalse) m-> (m~ p)].
\Proof. mv. intro p. intro H. intro H0. 
destruct (H H0) as [p0 [H1 H2]]. apply H2. exact H1. \Qed.
